---
WGCNA Fibroblasts in senescence
---

#Choose working directory

# Set up environment
```{r}
#clear workspace and set string as factors to false
rm(list=ls())
options(stringsAsFactors = F)

```


load packages
```{r}
library(openxlsx)
library(readxl)
library(WGCNA)
library(rstudioapi)
library(dplyr)
library(RCy3)
library(readr)
```

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

# read data and combine them to input file for WGCNA
```{r}
data <- read.xlsx("Merged_data.xlsx")
names(data)
rownames(data) <- data$Ensembl_GeneID
data <- data[, -c(1:6)]
# Obtén los nombres de las columnas que comienzan con "MRC"
selected_columns <- names(data)[grepl("^HFF", names(data))]

# Crea un nuevo data frame con solo las columnas seleccionadas
data <- data[, selected_columns]
head(data)

```

# check if there are samples with missing data
```{r}
gsg = goodSamplesGenes(data, verbose = 3);
gsg$allOK

# normalized counts from RNA-seq data should be log-transformed
data.log <- log2(data+1)
data.log$sum <- rowSums(data.log)
data.log <- data.log[data.log$sum > 0.5,]
data.log <- as.data.frame(t(data.log))
```


```{r}
traitData = read.table("Senescence_SampleData_.txt", header=TRUE);

data.filtered = data.log[row.names(data.log) %in% traitData$SampleID,]
colnames(data.filtered)

# Form a data frame analogous to expression data that will hold the clinical traits.
samples = rownames(data.filtered);
traitRows = match(samples, traitData$SampleID);
#datTraits = traitData[traitData$SampleID %in% samples, "Senescence", drop=FALSE]
datTraits = traitData[traitRows, -1];
rownames(datTraits) = traitData[traitData$SampleID %in% samples, "SampleID"]

datTraits$Senescence <- as.numeric(datTraits$Senescence)
datTraits$Cell_Type <- as.numeric(datTraits$Cell_Type)

collectGarbage();
```

#merge together the filtered table with the information from the dataset
```{r}
# Cluster samples
sampleTree = hclust(dist(data.filtered), method = "average")

# Convert traits to a color representation: white means low, red means high, grey means missing entry
traitColors = numbers2colors(datTraits, signed = FALSE);
sizeGrWindow(12,12)

# Plot the sample dendrogram and the colors underneath.
plotDendroAndColors(sampleTree, traitColors,
                    groupLabels = names(datTraits), cex.dendroLabels = 0.5, 
                    main = "Sample dendrogram and trait heatmap")
```

```{r}
save(data.filtered, datTraits, file = "WGCNA-input.RData")
```


#########################################
Network construction and module detection
#########################################

```{r}
# Allow multi-threading within WGCNA. This helps speed up certain calculations.
# At present this call is necessary for the code to work.
# Any error here may be ignored but you may want to update WGCNA if you see one.
# Caution: skip this line if you run RStudio or other third-party R environments. 
# See note above.
#enableWGCNAThreads()
# Load the data saved in the first part
lnames = load(file = "WGCNA-input.RData");
#The variable lnames contains the names of loaded variables.
lnames
```

```{r}
# Choose a set of soft-thresholding powers
powers = seq(1,15, by=2)

# Call the network topology analysis function
sft = pickSoftThreshold(data.filtered, powerVector = powers, verbose = 5)

save(sft, file = "WGCNA-sft.RData")

# Plot the results:
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;

# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2], xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n", main = paste("Scale independence"));

text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2], labels=powers,cex=cex1,col="red");

# this line corresponds to using an R^2 cut-off of h
abline(h=0.80,col="red")

# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5], xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n", main = paste("Mean connectivity"))

text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```

```{r}
# looking at both - soft threshold and mean connectivity 
# I decided to go with power 6 for this small example dataset
net = blockwiseModules(data.filtered, power = 6,
                       TOMType = "unsigned", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE,
                       saveTOMFileBase = "expTOM", 
                       verbose = 3)

save(net, file = "WGCNA-net.RData")

```


```{r}
# open a graphics window
sizeGrWindow(15, 9)
# Convert labels to colors for plotting
mergedColors = labels2colors(net$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
```

```{r}
moduleLabels = net$colors
moduleColors = labels2colors(net$colors)
table(moduleColors)
MEs = net$MEs;
geneTree = net$dendrograms[[1]];
save(MEs, moduleLabels, moduleColors, geneTree, file = "network-reconstruction.RData")
```



##########################################
Relate modules to external clinical traits
##########################################

```{r}
# Load the expression and trait data saved in the first part
load(file = "WGCNA-input.RData");

# Load network data saved in the second part.
#load(file = "network-reconstruction.RData");
```

```{r}
# Define numbers of genes and samples
nGenes = ncol(data.filtered);
nSamples = nrow(data.filtered);
# Recalculate MEs with color labels
MEs0 = moduleEigengenes(data.filtered, moduleColors)$eigengenes
MEs = orderMEs(MEs0)
moduleTraitCor = cor(MEs, datTraits, use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples);
```

```{r}
sizeGrWindow(20,20)
# Will display correlations and their p-values
textMatrix =  paste(signif(moduleTraitCor, 2), "\n(",
                           signif(moduleTraitPvalue, 1), ")", sep ="");
dim(textMatrix) = dim(moduleTraitCor)
par(mar = c(8, 8.5, 3, 3));

# Display the correlation values within a heatmap plot
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = names(datTraits),
               yLabels = names(MEs),
               ySymbols = names(MEs),
               colorLabels = FALSE,
               colors = blueWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.3,
               cex.lab.y = 0.5,
               zlim = c(-1,1),
               main = paste("Module-trait relationships"))
```



En la representación, si es neg, es que se va reduciendo, cuanto más severa pues más neg, y así.
```{r}
# Define variable time containing the time column of datTrait
Senescence = as.data.frame(datTraits$Senescence);
names(Senescence) = "Senescence"
# names (colors) of the modules
modNames = substring(names(MEs), 3)

geneModuleMembership = as.data.frame(cor(data.filtered, MEs, use = "p"));
MMPvalue = as.data.frame(corPvalueStudent(as.matrix(geneModuleMembership), nSamples));

names(geneModuleMembership) = paste("MM", modNames, sep="");
names(MMPvalue) = paste("p.MM", modNames, sep="");

geneTraitSignificance = as.data.frame(cor(data.filtered, Senescence, use = "p"));
GSPvalue = as.data.frame(corPvalueStudent(as.matrix(geneTraitSignificance), nSamples));

names(geneTraitSignificance) = paste("GS.", names(Senescence), sep="");
names(GSPvalue) = paste("p.GS.", names(Senescence), sep="");
names(geneTraitSignificance)
names(GSPvalue)
```

```{r}
modules = c("brown","red","black")
sizeGrWindow(9, 3);
  par(mfrow = c(1,3));
for(module in modules) {
  column = match(module, modNames);
  moduleGenes = moduleColors==module;
  
  verboseScatterplot(abs(geneModuleMembership[moduleGenes, column]),
                     abs(geneTraitSignificance[moduleGenes, 1]),
                     xlab = paste("Module membership (MM,", module, ")", sep=""),
                     ylab = "Gene significance (GS) for disease",
                     main = paste("MM vs. GS\n"),
                     cex.main = 1.2, cex.lab = 1.2, cex.axis = 1.2, col = module)
}
```

```{r}
# Change column names
names(geneTraitSignificance) = "GS_Senescence"
names(GSPvalue) = "p.GS_Senescence"


# Create the starting data frame
geneInfo0 = data.frame(Gene.ID = colnames(data.filtered),
                      moduleColor = moduleColors,
                      geneTraitSignificance,
                      GSPvalue)

# Order modules by their significance for time
modOrder = order(-abs(cor(MEs, Senescence, use = "p")));

# Add module membership information in the chosen order
for (mod in 1:ncol(geneModuleMembership))
{
  oldNames = names(geneInfo0)
  geneInfo0 = data.frame(geneInfo0, geneModuleMembership[, modOrder[mod]], 
                         MMPvalue[, modOrder[mod]]);
  names(geneInfo0) = c(oldNames, paste("MM.", modNames[modOrder[mod]], sep=""),
                       paste("p.MM.", modNames[modOrder[mod]], sep=""))
}
# Order the genes in the geneInfo variable first by module color, then by geneTraitSignificance
#"GS.senescence" %in% colnames(geneInfo0)

geneInfo0$GS_Senescence <- as.numeric(as.character(geneInfo0$GS_Senescence))

geneOrder = order(geneInfo0$moduleColor, -abs(geneInfo0$GS_Senescence));
geneInfo = geneInfo0[geneOrder, ]
```

```{r}
write.csv(geneInfo, file = "geneInfo.csv", row.names = FALSE)
```

